Helmet Foam Case Study Report

### 2024-05-07 CS2 Final Draft.qmd
Update CS2 Final Draft.qmd

---
title: "Helmet Foam Case Study Report"
author: 'Katie Greene, Ava Harman, Chris Seery, Ian Stonecypher'
date: 'May, 2024'
format: pdf
---

@@

```{r}
#| echo: false
#| label: fig-1
#| warning: false
#| fig-cap: "Histogram of Frequencies Tested"
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 4

@@

ggplot(foam_Samples, aes(x=amp)) +
  geom_histogram(bins=10)
```

The histogram in @fig-1 shows the distribution of compression amplitudes among the foam samples, revealing that certain amplitudes were tested much less frequently than others. These non-uniformities in sample distribution might impact the reliability of conclusions drawn from the analysis.

@@

Based on the research questions, we developed two separate models: one for stiffness fraction and the other for damping fraction. Due to the experimental design, and the fact that measurements from the same sample are not independent of each other, we used a mixed model format for our predictions. Log(cycle) was treated as a level-one variable because it changes across individual measurements, while all other variables were treated as level-two variables because they only vary across foam samples. In a mixed model, level-two variables impact the model intercept, and level-one variables interact with the level-two variables to effect the model slope. This accounts for each sample appeared to react to log(cycle) differently.

@@

$\beta_i = \beta_0  + \beta_2*\text{chemIndex} +\beta_3*\text{freq}$

@@

The summary of this model shows a high significance in almost all of our coefficients through small p-values (less than 0.05). These small p-values suggest that these variables are significant predictors of our response variable. Like the stiffness model, there are some predictors that do not have a significant p-value but their interaction terms are very significant. We can double check our predictors in the below table as well and see that all our predictors are significant.

@@

```{r, echo = FALSE}
#| echo: false
#| label: fig-7
#| warning: false
#| fig-cap: "Intercept Rand Effects vs. Slope Rand Effects"
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 4

@@

**Question 1:** We see that frequency, amplitude, chemical index, and cycle all have significant impacts on both stiffness and damping.

**Question 2:** For stiffness, we see that the individual predictors on their own will increase the initial stiffness, but as time goes on, foam using chemical index 121 will have the greatest decrease in stiffness. The individual predictors will impact damping in a similar way but foam with a chemical index of 79 will see the greatest decay in damping. Overall, there appear to be significant impacts on the stiffness and damping values depending on chemical index and frequency.

@@

A similar concern comes from foam types and testing conditions that were used less often, such as high amplitude The reliability of our models is already limited, and for these cases they are even less reliable. Additionally, if one variety of foam was never tested at high amplitude, then the model's claims about that foam at high frequency are based on nothing, and should not be trusted at all. So not all information this model gives on what effects stiffness and damping in foam is equally trustworthy.

@@

ggplot(foam_Samples, aes(x=amp)) +
  geom_histogram(bins=10)

@@

#StiffnessFrac Model
library(car)
suppressWarnings(FactorModel1 <- lmer(log(stiffnessFrac) ~ factor(chemIndex) + freq + amp +  log(cycle) +  factor(chemIndex):log(cycle) +  freq:log(cycle) + amp:log(cycle) + (1 + log(cycle)|ID), data = foam_data_scaled))
summary(FactorModel1)$coef
Anova(FactorModel)

# random effects of stiffnessFrac model
re <- ranef(FactorModel)$ID

plot(re)

@@

# random effects of dampingFrac model
re <- ranef(dampingModel)$ID

>>>>

---
title: "Helmet Foam Case Study Report"
author: 'Katie Greene, Ava Harman, Chris Seery, Ian Stonecypher'
date: '2024-05-06'
format: pdf
---

@@

```{r}
#| echo: false
#| label: fig-1
#| warning: false
#| fig-cap: "Histogram of Frequencies Tested"
#| fig-align: "center"
#| fig-width: 4
#| fig-height: 3

@@

ggplot(foam_Samples, aes(x=freq)) +
  geom_histogram(bins=10)
```

The histogram in @fig-1 shows the distribution of compression frequencies among the foam samples, revealing that certain frequencies were tested much less often than others. These non-uniformities in sample distribution might impact the reliability of conclusions drawn from the analysis.

@@

Based on the research questions, we developed two separate models: one for stiffness fraction and the other for damping fraction. Due to the experimental design, and the fact that measurements from the same sample are not independent of each other, we used a mixed model format for our predictions. Log(cycle) was treated as a level-one variable because it changes across individual measurements, while all other variables were treated as level-two variables because they only vary across foam samples. In a mixed model, level-two variables impact the model intercept, and level-one variables interact with the level-two variables to effect the model slope. This accounts for how each sample appeared to react to log(cycle) differently.

@@

$\beta_i = \beta_0  + \beta_1*\text{chemIndex} +\beta_2*\text{freq}$

@@

The summary of this model shows a high significance in almost all of our coefficients through small p-values (less than 0.05). These small p-values suggest that these variables are significant predictors of our response variable.

Like in the stiffnessFrac model, the interaction between log(cycle) and frequency has a negative coefficient, so damping decreases more at higher compression frequencies. Chemical index levels 2 and 3 (aka 100 and 121) interacted with log(cycle) also have negative coefficients, indicating that these foams lose damping faster than foam with index 79. But like in the stiffnessFrac model, these two chemical indices have higher damping right after compressions start, and foams with high frequency do as well.

Like the stiffness model, not all chemical indices look significant in Table 7, but we can double-check our predictors in the Anova in Table 8 and see that all our predictors are significant.

@@

```{r, echo = FALSE}
#| echo: false
#| label: fig-7
#| warning: false
#| fig-cap: "Intercept Rand Effects vs. Slope Rand Effects"
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 4

@@

**Question 1:** We see that frequency, chemical index, and cycle all have significant impacts on both stiffness and damping. Amplitude also has an impact on stiffness.

**Question 2:** For stiffness, we see that high frequency, high amplitude, and chemical indices 100 and 121 will increase the stiffness right after compressions start, but as time goes on, these will cause the more decrease in stiffness. Foam using chemical index 121 will have the greatest decrease in stiffness, and foam using chemical index 79 will have its stiffness decrease the least. For damping, high frequency and chemical indices 100 and 121 similarly lead to higher damping right after compressions start, and lead to the fastest decay in damping. The type of foam with the greatest decay in damping is chemical index 121, and foam using chemical index 79 will have its damping decrease the least.

@@

A similar concern comes from testing conditions that were used less often, such as high frequency. The reliability of our models is already limited, and for these cases they are even less reliable. Additionally, if one variety of foam was never tested at high amplitude, then the model's claims about that foam at high frequency are based on nothing, and should not be trusted at all. So not all information this model gives on what effects stiffness and damping in foam is equally trustworthy.

@@

ggplot(foam_Samples, aes(x=freq)) +
  geom_histogram(bins=10)

@@

#StiffnessFrac Model
library(car)
suppressWarnings(FactorModel1 <- lmer(log(stiffnessFrac) ~ factor(chemIndex) + freq + amp +  log(cycle) +  factor(chemIndex):log(cycle) +  freq:log(cycle) + amp:log(cycle) + (1 + log(cycle)|ID), data = foam_data_scaled))
summary(FactorModel1)$coef

Anova(FactorModel)

# random effects of stiffnessFrac model
re <- ranef(FactorModel)$ID
plot(re)

suppressWarnings(dampingModel <- lmer(log(stiffnessFrac) ~ factor(chemIndex) + freq + log(cycle) +  factor(chemIndex):log(cycle) + freq:log(cycle) + (1 + log(cycle)|ID), data = foam_data_scaled))
summary(dampingModel)$coef

@@

# random effects of dampingFrac model
re <- ranef(dampingModel)$ID
plot(re)
```
